Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-09 manuscript.tex
added Brenna fig, and reformatted

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent. \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

% add intro sentence here before jumping into data -- keep working on this
%AT-all responses are less clear given that there are no G:C base pairs for comparison.

In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response. In particular, the stretched fast response for AT-all and GC-end indicate that there may be a combination of processes occurring on the fast timescale. Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response. On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T vibration that points to A:T fraying as the predominant dynamic source. Given that fraying and shifting responses are not clearly distinguishable, we cannot determine whether AT-all and GC-end stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two. Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and increased stretching in the fast response. (Can briefly mention further experimental work to isolate shifting responses here?)

@@

\begin{figure} %[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Fig6.pdf}
        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events. Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs. The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined. Fluctuation between states is common during the transition.}
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

\subsection{Heterogeneous base pair distribution promotes canonical hybridization and dissociation}

@@

To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near hybridization and dissociation events, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}). During hybridization, we observed the formations of some key base pair contacts before the full duplex formed. Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T. This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}. For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs. There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs. In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or tICA space, nor do they tend to reform intact duplexes. As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}. We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=450, scale=1]{Fig7.pdf}
        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end. Color maps show inverse distances between out-of-register ends and complementary ends.}
        \label{fig:GC-end_dmaps}
	\end{center}
\end{figure}

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=300, scale=1]{Fig8.pdf}
        \caption{Sequence-dependent population distributions for trimmed hybridization events. $Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression. Populations are discretized in 200 bins and smoothed by averaging with the nearest 5 bins. Heat maps show the average intermolecular base distances in nm at increasing values of $Psi_2$, where the main diagonal shows native base pair distances.}
        \label{fig:Fig8}
	\end{center}
\end{figure}

@@

For each sequence, we compared population distributions along $\Psi_2$ and noted statistically meaningful differences in the range $ 0.001 <\Psi_2 < 0.0035$ (Figure \ref{fig:Fig8}). In this region, some base pair contacts have formed but frayed structures dominate the distribution. As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs. GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state. This is consistent with the downhill "nucleation-zippering" mechanism described above. Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et. al. regarding fdeviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}. This qualitative agreement shows that our MSM information can be adapted to identify dynamic behavior within the lag time and explore different definitions of two-state deviations.

@@

We have demonstrated how coarse-grained MD can be supplemented by data-driven time-lagged analysis to efficiently learn the slowest processes of DNA hybridization. We replicated T-jump analysis across a series of temperatures to verify previous spectroscopic assignments and sequence-dependent effects at different timescales. We constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent dynamics, and showed the relevance of these dynamics to experimental results. We found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all. On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted. This approach allowed us to aggregate an ensemble of trajectories in order to properly sample hybridization and dissociation without bearing a large computational expense.

Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons. These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=120mm, 
        scale=0.5]{Figs/figs_imp/GCSeq_TJTraces.png}
        \caption{Normalized time traces from T-jump HDVE data probing the change in in-plane nucleobase vibrations. The signal at 1600 cm-1 corresponds to A and T nucleobases while the signal at 1660 cm-1 contains G, C, and T contributions. Each T-jump was performed with a temperature change of ~15 C to a final temperature near the Tm of each respective sequence.}
        \label{fig:fast_stretching}
	\end{center}
\end{figure}

@@

%\bibliography{references}
\bibliography{refs_mendeley2}

>>>>

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

We initialized four self-complementary double-stranded sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent. \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

\begin{figure}[ht!] % right place for NN table? could go to SI
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{Fig6.pdf}
        \caption{Normalized time traces from T-jump HDVE data probing the change in in-plane nucleobase vibrations. The signal at 1600 cm-1 corresponds to A and T nucleobases while the signal at 1660 cm-1 contains G, C, and T contributions. Each T-jump was performed with a temperature change of 15 °C to a final temperature near the Tm of each respective sequence. Each time trace was fit to the sum of a stretched exponential and two exponentials (solid lines). The stretched exponential describes the component from 10 ns to 1 μs ($tau_{Fast}$), and the two exponentials describe the signal rise from 1 - 300 μs ($tau_{Slow}$) and thermal relaxation of the sample back to the initial temperature. The stretch factor (β) from each fit is listed.}

        \label{fig:Fig6}
	\end{center}
\end{figure}

In addition to identifying relaxation timescales, the T-jump IR response profile can inform the distribution of dynamics contributing to the response. In particular, the fast response of AT-all and GC-end exhibits more stretched exponential character than for GC-core and GC-mix  (Figure \ref{fig:Fig6}). Furthermore, the GC-end response contains both G:C and A:T vibrational contributions, suggesting that native terminal base fraying may not be the unique source of the response. On the other hand, GC-mix and GC-core are characterized by a distinct fast response comprised only of A:T features that points to A:T fraying as the predominant dynamic source. We predict that the mid-IR responses of fraying and shifting are not clearly distinguishable for GC-ends and AT-all, therefore we cannot determine the stretched responses originate from fraying dynamics in out-of-register configurations, shifting mechanisms during the dissociation process, or some combination of the two. Although reproducing stretched signals is not within the scope of this analysis, we see qualitative agreement between more complex transition network (AT-all > GC-end > GC-core ~ GC-mix) and increased stretching in the fast response.

@@

\begin{figure} %[ht!]
	\begin{center}
	\includegraphics[width=150mm, scale=1]{Fig7.pdf}
        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events. Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs. The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined. Fluctuation between states is common during the transition.}
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

\subsection{Heterogeneous base pair distribution promotes nucleation-zippering hybridization and dissociation}

@@

To supplement GC-mix analysis, we looked at qualitative trends in our trajectory data near hybridization and dissociation events, paying close attention to the distances between matching WC-pairs (Figure \ref{fig:GC-mix_transitions}). During hybridization, we observed the formations of some key base pair contacts before the full duplex formed. Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T. This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}. For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs. There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs. In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or tICA space, nor do they tend to reform intact duplexes. As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}. We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=150mm, scale=1]{Fig8.pdf}
        \caption{First two diffusion map coordinates built from 10000 5S1 states, equally sampled from AT-all and GC-end. Color maps show inverse distances (nm$^{-1}$) between out-of-register ends and complementary ends. The lower right region of the diffusion map space shows a high concentration of "shifted loop" base pairs that may facilitate the out-of-register GC-end transition.}
        \label{fig:GC-end_dmaps}
	\end{center}
\end{figure}

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=150mm, scale=1]{Fig9.pdf}
        \caption{Sequence-dependent population distributions for trimmed hybridization events. $\Psi_2$ represents the maximum structural variance across all sequence configurations and correlates strongly with hybridization progression. Populations are discretized in 200 bins and smoothed by averaging with the nearest 5 bins. Heat maps show the average intermolecular base distances in nm at increasing values of $\Psi_2$, where the main diagonal shows native base pair distances.}
        \label{fig:Fig8}
	\end{center}
\end{figure}

@@

Once we determined a strong hybridization coordinate, we compared probability distributions for each sequence along $\Psi_2$ (Figure \ref{fig:Fig8}). The $ <\Psi_2 < 0$ reflects the distance between approaching strands, but we noted statistically meaningful differences between sequence once first WC contacts are made. In the $ 0.001 <\Psi_2 < 0.0035$ region, heatmaps show that contacts tend to form near the center of the duplex, and frayed structures progressively zipper closed as $<\Psi_2$ increases. As expected, GC-core shows a distinct peak in the frayed region, corresponding to the F4 macrostate identified by SRV-MSMs. GC-mix has a notably higher population than AT-all and GC-end but displays a more gradual progression toward the native hybridized state. This is consistent with the downhill nucleation-zippering mechanism described above. Although differences between AT-all and GC-end are more subtle, overall sequence populations in this region follow predictions by Sanstead et. al. regarding deviations from two-state binding behavior, where GC-core > GC-mix > AT-all > GC-end \citep{Sanstead2016}. This qualitative agreement shows that MSM information can be adapted to compare sequences along common pathways and elucidate dynamics occurring faster than the lag time. Furthermore, we are able to distinguish out-of-register deviations from fraying deviations and describe relevant experimental comparisons for each.

@@

We have demonstrated how G:C placement in an otherwise repetitive AT sequence has a profound impact on equilibrium hybridization dynamics. By supplementing coarse-grained MD with data-driven time-lagged analysis, we constructed high resolution SRV-MSMs capable of distinguishing sequence-dependent kinetic behavior. In particular, we found that AT-all and GC-end sequences both participate in some degree of out-of-register base pair shifting, although these states have higher kinetic relevance for AT-all. On the other hand, GC-core hybridization transits through, or is perhaps facilitated by, a frayed intermediate in which one half of A:T bonds are broken and the duplex is significantly disrupted. Our computational approach and results show strong comparisons T-jump experiments at ns and us timescales. Going forward, we expect that this work will be extended to predict more general trends in sequence-dependent hybridization and motivate strategies for experimental comparisons. These insights should be leveraged for sequence design and incorporated into the growing field of dynamic DNA nanotechnology.

@@

@@

\bibliography{references}
%\bibliography{refs_mendeley2}
